Bayesian model comparison for rare variant association studies of multiple phenotypes

### 2017-12-22 tex/MVMetaLearnMRP2017OctoberMRPMM.tex

\bibliographystyle{naturemag}
\title{Bayesian model comparison for rare variant association studies of multiple phenotypes}
\author{Christopher DeBoever$^{1}$, Matthew Aguirre$^{1}$, Yosuke Tanigawa$^{1}$, Adam Lavertu$^{1}$, Greg McInnes$^{1}$, Timothy Poterba$^{2}$, Mark J. Daly$^{2,3}$,  Matti Pirinen$^{4}$\thanks{matti.pirinen@helsinki.fi}, and Manuel A. Rivas$^{1}$\thanks{mrivas@stanford.edu}}
\usepackage{setspace}   
\linenumbers
\begin{document}
\maketitle
\begin{abstract}
\singlespacing
Whole genome sequencing studies applied to large populations or biobanks with extensive phenotyping raise new analytic challenges. The need to consider many variants at a locus or group of genes simultaneously and the potential to study many correlated phenotypes with shared genetic architecture provide opportunities for discovery and inference that are not addressed by the traditional one variant-one phenotype association study. Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies. In so doing we consider the use of summary statistic data to 1) apply standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, which may be used in practice for downstream quality control, and 3) models for identifying protective protein-truncating variants, which can expedite drug discovery.  Through simulation studies, we demonstrate that the proposed model comparison approach may improve ability to detect rare variant association signals. Finally, we demonstrate its use with rare variant data combined with asthma diagnosis and haematological and spirometry measures for individuals in the UK Biobank. We show that we are able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.
\end{abstract}

@@

\section{Introduction} 
%% Sequencing technologies
Sequencing technologies are quickly transforming human genetic studies of complex traits: it is increasingly possible to obtain whole genome sequence data on thousands of samples at manageable costs. As a result, the genome-wide study of rare variants (minor allele frequency [MAF] $< 1\%$) and their contribution to disease susceptibility and phenotype variation is now feasible\cite{ifih1,altshuler2010map,rivas2011deep,10002012integrated}. 
%% Rare variant methods
In genetic studies of diseases or continuous phenotypes, rare variants are hard to assess individually due to the limited number of copies. Hence, to boost the ability to detect a signal, evidence is usually `aggregated' across variants. When designing an `aggregation' method, there are three questions that are usually considered. First, across which biological units should variants be combined; second, which variants mapping within those units should be included\cite{majithia2014rare}; and third, which statistical model should be used\cite{lee2014rare}. Given the widespread observations of shared genetic risk factors across distinct diseases, there is also considerable motivation to proactively modify gene discovery approaches accordingly. In other words, rather than only aggregating variants that may have effects on a single phenotype, we can also bring together sets of phenotypes for which a single variant or sets of variants might have effects.

@@

%% Use of summary statistic data
In practice, sharing of individual genotype and phenotype data across groups in large genetic consortium studies is difficult to achieve.  We present a derivation where summary statistics, such as estimates of effect size and the corresponding standard error from single variant and single phenotype analysis, are the input data. Furthermore, we use insights from Liu et al.\cite{liu2014meta} and Cichonska et al.\cite{cichonska2016metacca} who suggest the use of additional summary statistics, like covariance estimates across variants and studies, respectively, that would enable lossless ability to detect gene-based association signals using summary statistics alone.

%% Genome variant annotation and protein truncating variants
Genome variant annotation is critical for the interpretation of genetic findings. In the MRP model comparison we present approaches for including priors on spread of effect sizes. An example of an important class of variants where functional prediction is much more straightforward is the class of variants predicted to truncate the protein product, commonly referred to as protein truncating variants or PTV\cite{rivas2013assessing,rivas2015effect}, and in this paper we place special emphasis on the interpretation of PTVs.

@@

%%Apply methods to the GLGC data and exome sequencing data sets. 
We apply MRP to four phenotypes: asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV1, INI3063), and Forced Vital Capacity (FVC, INI3062) in the UK Biobank dataset. We show that we are able to retain useful features from single variant and single phenotype meta-analysis, rare variant analysis, and prioritize protective modifiers of disease risk.

%% FIGURE 1 
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{../figures/FIG1FINAL.pdf}
  \caption{Schematic overview of MRP. A. MRP is suitable for a broad range of rare variant association study designs including (from left to right): i) case-control, ii) multiple diseases with shared controls, iii) single quantitative phenotype, and iv) mixture of case-control and quantitative phenotypes. B. Diagram of factors considered in rare variant association analysis including the correlation matrices: $\mathbf{R_{\textrm{study}}}$  (expected correlation of genetic effects among a group of studies), $\mathbf{R_{\textrm{var}}}$ (expected correlation of genetic effects among a group of variants), and $\mathbf{R_{\textrm{phen}}}$ (expected correlation of genetic effects among a group of phenotypes); the scale parameter for genetic variant annotation; and the location of genetic effects, which may be used to prioritize or identify protective modifiers of disease risk.}
  \label{Fig1}
  \end{figure}

@@

Let $N$ be the number of individuals and $K$ the number of quantitative phenotype measurements on each individual, for example, here in the applications we focus on asthma, eosinophil count, Forced Expiratory Volume in 1-second, and Forced Vital Capacity, in which case $K=4$. Let $M$ be the number of variants in a testing unit ${\bs G}$, where ${\bs G}$ can be, for example, a gene, pathway, or a network. Let $S$ be the number of studies where data is obtained from - this data may be in the form of raw genotypes and phenotypes or summary statistics including linkage-disequilibrium, effect sizes (or odds ratio), and standard error of the effect size.

@@

For gene discovery, the scale may be used to denote the prior spread of effect sizes. For instance, emerging empirical genetic studies have shown that within a gene protein truncating variants may have stronger effects than missense variants\cite{do2015exome}.

@@

\newpage 
%% FIGURE 2
\begin{figure}
\centering
   \includegraphics[width=\textwidth]{../figures/Figure2Nov2017.pdf}
  \caption{Simulation studies. A. Comparison of $-\log_{10}$(p-values) from frequentist $\textrm{BF}_{\textrm{MRP}}$ approximation for an independent effects and a similar effects model to commonly used gene-based statistical tests (skat and burden). B. Comparison of log10(Bayes Factors) obtained when raw genotype and phenotype data is available to a scenario where summary statistics only was available and similar effects across studies is assumed. C. From single variant and single phenotype to multiple variants and multiple phenotypes gene discovery: ROC curves for detecting gene association to any of the phenotypes using single variant/single phenotype association (blue) to multiple variants and multiple phenotypes association (purple). D. ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles. }
  \label{Fig2}
\end{figure}

@@

%% FIGURE 3
\begin{figure}
\centering
\includegraphics[width=\textwidth]{../figures/manhattan.pdf}
  \caption{MRP application to summary statistics from (A) 
  MRP application to summary statistic data from the GLGC and EOMI exome-chip studies. (A-E). Here, we show manhattan plots of meta-analysis results obtained by applying gene-based MRP across PTVs for (A) triglycerides (TG), (B) low-density lipoprotein-cholesterol (LDL-C), (C) high-density lipoprotein cholesterol (HDL-C), and (D) EOMI separately. (E-F) Gene-based MRP results when multiple phenotypes are analyzed and prior mean of genetic effect is included to prioritize protective and risk signals. For each gene the the $\log_{10}\left(\textrm{Bayes Factor}\right)$ is plotted.}
  \label{Fig3}
\end{figure}

>>>>

\bibliographystyle{naturemag}
\title{Bayesian model comparison for rare variant association studies of multiple phenotypes}
\author{Chris Mark DeBoever$^{1}$, Matthew Aguirre$^{1}$, Yosuke Tanigawa$^{1}$, Adam Lavertu$^{1}$, Greg McInnes$^{1}$, Timothy Poterba$^{2}$, Mark J. Daly$^{2,3}$,  Matti Pirinen$^{4}$\thanks{matti.pirinen@helsinki.fi}, and Manuel A. Rivas$^{1}$\thanks{mrivas@stanford.edu}}
\usepackage{setspace}   
\linenumbers
\begin{document}
\maketitle
\begin{abstract}
\singlespacing
Whole genome sequencing studies applied to large populations or biobanks with extensive phenotyping raise new analytic challenges. The need to consider many variants at a locus or group of genes simultaneously and the potential to study many correlated phenotypes with shared genetic architecture provide opportunities for discovery and inference that are not addressed by the traditional one variant-one phenotype association study. Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies. In so doing we consider the use of summary statistic data to 1) apply standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, which may be used in practice for downstream quality control, and 3) models for identifying protective protein-truncating variants, which can expedite drug discovery.  Through simulation studies, we demonstrate that the proposed model comparison approach may improve ability to detect rare variant association signals. Finally, we demonstrate its use with rare variant data and haematological measures in individuals from multiple population groups in the UK Biobank. We show that we are able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.
\end{abstract}

@@

\section{Introduction} 
%% Sequencing technologies
Sequencing technologies are quickly transforming human genetic studies of complex traits: it is increasingly possible to obtain whole genome sequences on thousands of samples at manageable costs. As a result, the genome-wide study of rare variants (minor allele frequency [MAF] $< 1\%$) and their contribution to disease susceptibility and phenotype variation is now feasible\cite{ifih1,altshuler2010map,rivas2011deep,10002012integrated}. 
%% Rare variant methods
In genetic studies of diseases or continuous phenotypes, rare variants are hard to assess individually due to the limited number of copies. Hence, to boost the ability to detect a signal, evidence is usually `aggregated' across variants. When designing an `aggregation' method, there are three questions that are usually considered. First, across which biological units should variants be combined; second, which variants mapping within those units should be included\cite{majithia2014rare}; and third, which statistical model should be used\cite{lee2014rare}. Given the widespread observations of shared genetic risk factors across distinct diseases, there is also considerable motivation to proactively modify gene discovery approaches accordingly. In other words, rather than only bringing together variants that may have effects on a single phenotype also bring together sets of phenotypes for which a single variant or sets of variants might have effects.

@@

%% Use of summary statistic data
In practice, sharing of individual genotype and phenotype data across groups in large genetic consortium studies is difficult to achieve.  We present a derivation where summary statistics, such as estimates of effect size and standard error of a single variant and single phenotype analysis, are the input data. Furthermore, we use insights from Liu et al.\cite{liu2014meta} and Cichonska et al.\cite{cichonska2016metacca} who suggest the use of additional summary statistics, like covariance estimates across variants and studies, respectively, that would enable lossless ability to detect gene-based association signals using summary statistics alone.

%% Genome variant annotation and protein truncating variants
Genome variant annotation is critical for the interpretation of genetic findings. In MRP model comparison we present approaches for including priors on spread of effect sizes. An example of an important class of variants where functional prediction is much more straightforward is that which is predicted to truncate the protein product, commonly referred to as protein truncating variants or PTV\cite{rivas2013assessing,rivas2015effect}, and in this paper we place special emphasis on the interpretation of PTVs.

@@

%%Apply methods to the GLGC data and exome sequencing data sets. 
We apply MRP to four phenotypes: asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), Eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV1, INI3063), and Forced Vital Capacity (FVC, INI3062) in the UK Biobank dataset. We show that we are able to retain useful features from single variant and single phenotype meta-analysis, rare variant analysis, and prioritize protective modifiers of disease risk.

%% FIGURE 1 
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{/Users/mrivas/Dropbox/MultiplePhenotypesManuscript/figure/FIG1FINAL.pdf}
  \caption{Schematic overview of MRP. A. MRP is suitable for a broad range of rare variant association study designs including (from left to right): i) case-control, ii) multiple diseases with shared controls, iii) single quantitative phenotype, and iv) mixture of case-control and quantitative phenotypes. B. Diagram of factors considered in rare variant association analysis including the correlation matrices: $\mathbf{R_{\textrm{study}}}$  (expected correlation of genetic effects among a group of studies), $\mathbf{R_{\textrm{var}}}$ (expected correlation of genetic effects among a group of variants), and $\mathbf{R_{\textrm{phen}}}$ (expected correlation of genetic effects among a group of phenotypes); the scale parameter for genetic variant annotation; and the location of genetic effects, which may be used to prioritize or identify protective modifiers of disease risk.}
  \label{Fig1}
  \end{figure}

@@

Let $N$ be the number of individuals and $K$ the number of quantitative phenotype measurements on each individual, for example, here in the applications we focus on asthma, Eosinophil count, Forced Expiratory Volume in 1-second, and Forced Vital Capacity, in which case $K=4$. Let $M$ be the number of variants in a testing unit ${\bs G}$, where ${\bs G}$ can be, for example, a gene, pathway, or a network. Let $S$ be the number of studies where data is obtained from - this data may be in the form of raw genotypes and phenotypes or summary statistics including linkage-disequilibrium, effect sizes (or odds ratio), and standard error of the effect size.

@@

For gene discovery, the scale, may be used to denote the prior spread of effect sizes. For instance, emerging empirical genetic studies have shown that with a gene protein truncating variants may have stronger effects than missense variants\cite{do2015exome}.

@@

\newpage 
%% FIGURE 2
\begin{figure}
\centering
   \includegraphics[width=\textwidth]{/Users/mrivas/Dropbox/MultiplePhenotypesManuscript/Figure2Nov2017.pdf}
  \caption{Simulation studies. A. Comparison of $-\log_{10}$(p-values) from frequentist $\textrm{BF}_{\textrm{MRP}}$ approximation for an independent effects and a similar effects model to commonly used gene-based statistical tests (skat and burden). B. Comparison of log10(Bayes Factors) obtained when raw genotype and phenotype data is available to a scenario where summary statistics only was available and similar effects across studies is assumed. C. From single variant and single phenotype to multiple variants and multiple phenotypes gene discovery: ROC curves for detecting gene association to any of the phenotypes using single variant/single phenotype association (blue) to multiple variants and multiple phenotypes association (purple). D. ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles. }
  \label{Fig2}
\end{figure}

@@

%% FIGURE 3
\begin{figure}
\centering
\includegraphics[width=\textwidth]{/Users/mrivas/git/mrp/figures/manhattan.pdf}
  \caption{MRP application to summary statistic data from the GLGC and EOMI exome-chip studies. (A-E). Here, we show manhattan plots of meta-analysis results obtained by applying gene-based MRP across PTVs for (A) triglycerides (TG), (B) low-density lipoprotein-cholesterol (LDL-C), (C) high-density lipoprotein cholesterol (HDL-C), and (D) EOMI separately. (E-F) Gene-based MRP results when multiple phenotypes are analyzed and prior mean of genetic effect is included to prioritize protective and risk signals. For each gene the the $\log_{10}\left(\textrm{Bayes Factor}\right)$ is plotted.}
  \label{Fig3}
\end{figure}
